Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-04-28 doc/network.tex
going to work on it again tomorrow...

The individual-level method assumes that the initial exponential growth phase of an epidemic can be approximated by a nonhomogeneous Poisson process.

to derive a likelihood for the observations from each individual:
\begin{equation}
\begin{aligned}
&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta) \right),
\end{aligned}
\end{equation}
where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by infector $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
\cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
\cite{trapman2016inferring} estimated the true reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
however, individuals in later generations are potentially subject to stronger local depletion and even global depletion effects.
Therefore, we expect our empirical estimate to better reflect the infection process during the exponential growth phase.
\swp{TODO: Compare Trapman vs our estimate of empirical R0 in Appendix; Trapman estimate is much lower.}

>>>>

\swp{Cite something about poisson process}
We refer to the second method as the individual-level method because it relies on individual contact information.
We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate is equal to the effective generation-interval distribution.
Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is proportional to
\begin{equation}
\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta),
\end{equation}
where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
We use gamma distribution to model the effective generation-interval distribution;
parameters $\theta$ correspond to the mean and shape parameters of a gamma distribution.

The entire data likelihood of the individual-level method can be written as:
\begin{equation}
\begin{aligned}
&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta) \right),
\end{aligned}
\end{equation}
where $N$ is the total number of infected individuals.
Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
\cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.

\swp{Probabaly move simulations to appendix}
To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
\cite{trapman2016inferring} estimated the true reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
however, individuals in later generations are potentially subject to stronger local depletion and even global depletion effects.
Therefore, we expect our empirical estimate to better reflect the infection process during the exponential growth phase.
\swp{TODO: Compare Trapman vs our estimate of empirical R0 in Appendix; Trapman estimate is much lower.}
